---
title: "Base R Descriptive Stats"
author: "Kim Julia Fuchs"
date: "20 3 2020"
output: word_document
---
# libraries
```{r}
library(tidyverse)
```
# remove objects: rm()
```{r}
# remove specific objects by simply listing them 
# rm(<object1>, <object2> (...))
```
# list all variable names in given environment 
```{r}
ls()
```
# Summary stats
## summary(x): output Min, 1. Quartile, Median, 3. Quartile und Max
## mean(, na.rm = F): mean
## median(, na.rm = F): median
## var(,na.rm = F: variance
## sd(,na.rm = F): standard deviation
## range(,na.rm = F): returns min ([1]) and max ([2]) value of a vector

##  cov(x,y): covariance
##  cor(x, y = NULL, use = "everything", method = c("pearson", "kendall", "spearman")): correlation coeff.
  + x: numeric vector, matrix or data frame
  + w/ 'use' parameter one can handle missing values 
  + 'everything': NAs will propagate conceptually, i.e. resulting value will be NA whenever there is one NA
  + 'all obs.': presense of missing observation will result in error
  + 'complete obs.': casewise deletion of NAs, if no complete cases 'error'
  + 'na.or.complete': casewise deletion of NAs, if no complete cases 'NA'
  + 'pairwise.complete.obs'correlation or covariance between each pair of variables is computed (x must be matrix of df then) sing all complete pairs of observations on those variables
  
# Vectors: specifics 
## cumsum(): **Vectorized** cumulative sum
```{r}
x <- c(1:3)
cumsum(x)
```
## prod(): Product over elements within vector: 
* syntax: prod(..., na.rm = FALSE)
  + numeric, logical or complex vector
```{r}
prod(x)
```
## diff(): **Vectorized** difference within numeric vector or matrix
```{r}
x <- c(1, 1, 2, 6)
diff(x) # 0 1 4
```
# Matrices: specifics

* the following functions - colMeans, colSums, rowMeans and rowSums - are equivalent to apply() with FUN as sum or mean respectively
* **however, they are a lot faster**
 
## colMeans(): Arithmetic mean of each column
*syntax (same applies to following functions): colMeans(x, na.rm = FALSE, dims = 1)
  + x: matrix, array (of more than 2 dim), or numeric df
  + dims: integer; which dimensions are regarded as ‘rows’ or ‘columns’ to calculate over
  + na.rm: omitting/no omitting of NAs
    + note: omitting has the effect that mean may not be calculated over same number of rows or col respectively
    
## colSums(): Column-wise sums
## rowMeans(): Row-wise means
## rowSums(): Row-wise sums
```{r}
d <- data.frame(a = c(1,2,4,4), b = c(2,3,3,4))
d
colMeans(d)
colSums(d)
rowSums(d)
rowMeans(d)
```
## dim(): Dimensions of matrix: 
* returns nrow and ncol
```{r}
dim(d)
```
## nrow(): Number of rows 
## ncol(): Number of col
```{r}
nrow(d)
ncol(d)
```
# Data frames: specifics
* because dfs are a list of vector of equal lengths, they share properties of both, matrices and lists
* we therefore, can use functions for matrices and lists also for dataframes


# polyroot(): find roots (Nullstellen) of a real or complex polynomial
```{r}
# input: 
# vector of polynomial coefficients in increasing order 
# e.g. f(x) = x² + x + 1
polyroot(c(1, 1, 1))
# note: polyroot outputs complex values by default; to get real values use Re()
Re(polyroot(c(1,1,1)))
```
# uniroot(): find one (!) roots for any function in R
* as said, uniroot() can only finde one unit root 
* i.e. if function is not strictly concave/convex 
* input: function, interval in which to search for uniroot
```{r}
eq <- function(x) {
  x^2 - exp(x)
}

uniroot(eq)# error: 'interval missing
uniroot(function(x) x^2 - exp(x)) # error: 'interval missing'

# -> one has to check interval for which to calculate uniroot w/ plot first
# in the following I experiment a little w/ base R's visualizations
plot(eq, to = 2, from = -2, type = 'l',main = 'x^2 - exp(x)', ylab = 'f(x)')
# highlight a specific value by adding layer points()
points(x = c(uniroot(eq, lower = -1, upper = 2)$root),y = c(0), pch = 16, cex =2, col = 'green') 
# pch: plotting ‘character’, i.e., symbol to use. This can either be a single character or an integer code for one of a set of graphics symbols. The full set of S symbols is available with pch = 0:18, see the examples below. (NB: R uses circles instead of the octagons used in S.)

# add a line with certain values using abline()
abline(h = 0, col = 'red') # h = horizontal
abline(v= uniroot(eq, lower = -1, upper = 2)$root, col = 'blue') # v= vertical

# alternative: curve
curve(eq, to = 2, from = -2, type = 'l',main = 'x^2 - exp(x)', ylab = 'f(x)')
```

# Derivatives: D()
* syntax: D(expr, name)
* expr will be specific function
* name will be variable to derive by (df(x)/dx)

```{r}
D(quote(x^2 +3), 'x')
# Thanks to the quote() function, it keeps the expression unevaluated so that the symbols
# are directly accessible as they are written
```
# Integration: integrate()
* syntax: D(expr, lower bound, upper bound)
```{r}
integrate(function (x) cos (x), 0, pi/2)
```
# Sampling from a vector or list: sample()
*syntax: sample(x, size, replace = FALSE, prob = NULL)
* x = vector or list 
* size = number of items to choose  
* replace = should sampling be with replacement?
* prob = argument can be used to give a vector of weights for obtaining the elements of the vector being sampled. They need not sum to one
 + if replace is false, these probabilities are applied sequentially, i.e. prob. of choosing next item is proportional to weights amongst the remaining items 

# Random distribution: numeric simulations
* numeric simulation, i.e. drawing a sample from a random distribution 
* run ?Distributions to see a full list of supported distributions 
```{r}
# continuous uniform distribution: runif(n [numbers to be randomly drawn], min =  0 <or other lower bound>, max = 1 <or any other upper bound>)
?Distributions
runif(10, min = 0, max = 10)
plot(runif(1000, min = 0, max = 1))
hist(runif(1000, min=0, max =1))

# normal distribution: 
# rnorm(n [numbers to be randomly drawn], mean = 0 <or other mean value>, sd = 1 <or other standard deviation>)
rnorm(1000, mean = 0, sd =1)
plot(rnorm(1000, mean = 0, sd =1))
hist(rnorm(1000, mean = 0, sd = 1))
plot(dnorm(x = c(-10:10), mean = 0, sd = 1))
```

